A family of accessibility measures derived from spatial interaction principles

### 2025-10-23 paper_plos/paper_plos_V3.tex
cdh final proofread for v3

\section*{Abstract}
Transportation planning has long prioritized the efficiency of movement.
However, the concept of accessibility represents a more comprehensive
evolution, shifting focus from movement (i.e., trips) to the potential
to spatially interact with desired destinations. Despite growing
recognition of accessibility-based planning approaches, the concept
remains fragmented, with inconsistent definitions and unclear
interpretations. To this end, this paper offers a methodological
contribution by specifying a family of accessibility measures that are
grounded in the shared `gravity-based' theoretical roots of spatial
interaction models, particularly their balancing factors. From this
foundation, we outline four memebers of the family: the `unconstrained'
measure (i.e., Hansen-type accessibility), the `total constrained'
measure (i.e., a constrained version of the Hansen-type accessibility),
the `singly constrained' measure (i.e., related to the popular two step
floating catchment approach -- 2SFCA), and the `doubly constrained'
measure representing realized access (i.e., equal to the doubly
constrained spatial interaction model). These measures can be
interpreted as either the number of accessible opportunities or
accessible population (i.e., market potential). A toy example
illustrates how they produce interpretable unit-based values, offering a
clearer and more coherent basis for accessibility analysis.

\linenumbers
\section{1. Introduction}\label{introduction}

@@

This paper makes two contributions. First, we review how spatial
interaction modeling and accessibility share similar origins but have
diverged in focus and interpretation. Second, we introduce a family of
accessibility measures grounded in spatial interaction principles,
including total, singly, and doubly constrained cases and variants of
``accessible opportunities'' and ``accessible population''. These cases
and variants align with common measures such as Hansen-type
accessibility \citep{hansen1959}, competitive accessibility measures
such as the 2SFCA method \citep{shen1998, luo2003}, and market potential
models
\citep{harris_market_1954, vickermanAccessibilityAttractionPotential1974}.

@@

The remainder of this paper proceeds as follows. Sections 2 through 4
trace the historical development of spatial interaction and
accessibility research, beginning with Newtonian gravitational analogies
and Carey (1858) \citep{careyPrinciplesSocialScience1858} (Section 2),
moving through early researchers like Ravenstein (1885)
\citep{ravensteinLawsMigration1885} to Stewart (1948)
\citep{stewartDemographicGravitationEvidence1948} who theorised and
formalized spatial interaction patterns (Section 3), and examining how
the ``gravity-based'' accessibility approach in Hansen (1959)
\citep{hansen1959} became the dominant approach in planning practice
(Section 4). Section 5 presents the entropy-based family of spatial
interaction models in Wilson (1971) \citep{wilson1971}, explaining how
the introduction of ``constraints'' (based on top-down known information
as part of entropy-maximisation) produce interpretable, unit-consistent
flow estimates. Section 6 explores why accessibility and spatial
interaction literature diverged into separate branches despite their
shared conceptual foundations. Section 7 constitutes the paper's core
contribution: we derive a family of accessibility measures corresponding
to different constraints applied -- unconstrained, singly constrained,
and doubly constrained cases -- demonstrating how each yields zone-level
accessibility values expressed in meaningful units (opportunities or
population) rather than unit-inconsistent indices. We illustrate each
measure with numerical examples that clarify the practical implications
of different constraint assumptions. Section 8 concludes by discussing
how these constrained measures can inform planning decisions and improve
clarity in accessibility analysis.

@@

Where \(G\) is the gravitational constant -- an empirically calibrated
value that ensures the model reflects observed forces. Newton's initial
estimate of \(G\) was based on a speculation but received empirical
support after Hutton's and Cavendish' experiments in the late 1700s
\citep{hutton_xxxiii_1778, cavendish_xxi_1798} which estimated it to
within 1\% accuracy. That is, it took over a century from the
publication of Newton's \emph{Principia Mathematica} to refine the
estimate of the proportionality constant.

@@

Henry C. Carey's \emph{Principles of Social Science}
\citep{careyPrinciplesSocialScience1858} inspired empirical spatial
interaction research in different contexts. Namely, a number of
researchers theoretically and empirically attempted to characterise
human spatial interaction as a force \(F\) directly proportional to the
``masses'' \(M_i\) and \(M_j\) of two locations, and inversely
proportional to their separation distance -- conceptually parallel to
Newtonian gravity, but often omitting a proportionality constant.

@@

What is notable about Eq~\ref{eq-stewart-force}, however, is the
proportionality constant \(G\) was specified but ``left for future
determination''
\citep[p.~34]{stewartDemographicGravitationEvidence1948}. We can infer
that it is crucial for ensuring \(F\) is maintained in some units of
demographic force.

@@

From Stewart \citep{stewartDemographicGravitationEvidence1948}, we
arrive to 1959 and Walter G. Hansen, whose work proved to be
exceptionally influential in the accessibility literature
\citep{hansen1959}. In this seminal paper, Hansen defined accessibility
as the potential of opportunities for interaction\ldots{} a
generalization of the population-over-distance relationship or
\emph{population potential} concept developed by Stewart
\citep{stewartDemographicGravitationEvidence1948}'' (p.~73). As well as
being a student of city and regional planning at the Massachusetts
Institute of Technology, Hansen was also an engineer with the Bureau of
Roads, and preoccupied with the power of transportation to shape land
uses in a very practical sense. Hansen \citep{hansen1959} drew directly
from Stewart's population potential formula (see
Eq~\ref{eq-stewart-population-potential-sum}), but left aside the
broader (and often problematic) framework of ``social physics''.

@@

\(S_{i}\) in Eq~\ref{eq-accessibility-general} is a measure of the
accessibility from a zone \(i\). This is a function of \(O_j\) (the mass
of opportunities at \(j\)), \(d_{ij}\) (the cost, e.g., distance or
travel time, incurred to reach \(j\) from \(i\)), and \(f(d_{ij})\) (a
function that modulates the friction of cost). Today, Hansen is
frequently cited as the father of modern accessibility analysis
\citep[e.g.,][]{reggianiGuestEditorialNew2011}, and Hansen-type
accessibility is commonly referred to as the gravity-based accessibility
measure.

However, Hansen's formulation carried forward a crucial omission that
continues to affect the literature: the proportionality constant \(G\)
included in Stewart's original formulation
(Eq~\ref{eq-stewart-population-potential-sum}) has vanished entirely.
Although Stewart included \(G\) explicitly (with a note that ``\(G\)
{[}was{]} left for future determination: a suitable choice of other
units can reduce it to unity'' {[}p.~34{]}.) Hansen made no mention of
it. As a result, modern accessibility analysis has largely evolved
without addressing the constant's role, leaving \(G\) effectively and
implicitly fixed at 1. This omission has significant implications.
Without a proportionality constant, the accessibility formula expresses
only a proportional relationship:
\(S_i \propto \sum_j g(O_j)f(d_{ij})\), not one of calibrated equality.
Recognition of the nature of this relationship is not common in the
literature, but is known, i.e., this proportional equation is shown in
Figure 1 in Wu and Levinson \citep{wuUnifyingAccess2020}.

Furthermore, working with a proportional relationship generates
fundamental issues in comparability between and, arguably within,
studies. Namely, accessibility estimates have no fixed unit, rendering
them sensitive to the choice of impedance functions. For instance, if
travel cost \(d_{ij}\) is measured in meters, then when the travel
impedance function \(f(d_{ij})\) equals \(d_{ij}^{-\beta}\), the
resulting \(S_i\) has units of opportunities per
\(\text{metres}^{\beta}\). However, when \(f(d_{ij})\) is set to equal
\(e^{-\beta d_{ij}}\), the units become opportunities per
\(e^{\beta \text{metres}}\). Such variation impairs comparability across
analysis and obscures the meaning of accessibility scores, making them
difficult to understand and communicate without post-hoc treatment.

Therefore in practice, Hansen-type measures are ones of proportionality
and are better understood as \emph{ordinal indicators}; they rank
accessibility but lack cardinal meaning or consistent units
\citep{millerAccessibilityMeasurementApplication2018}. The continued
absence of a sort of proportionality constant \(G\) leaves a conceptual
and practical gap in accessibility analysis: a missing link between
theoretical form and empirical measurement that hinders the
interpretability of accessibility measures.

@@

While accessibility research evolved in North America with Hansen
\citep{hansen1959}, a parallel development was taking place across the
Atlantic with Alan G. Wilson. Wilson's ground breaking paper
\citep{wilson1971} advanced a general framework for spatial interaction
modeling focused on flows of interaction between places and derived from
a non-Newtonian analogy. Additionally, this work was not focused on the
`potential' concept as accessibility had been. Wilson \citep{wilson1971}
formalized the general spatial interaction model through the following
equation:

@@

Of important note for this paper, \(k\) in
Eq~\ref{eq-phys-gravity-model} acts as a proportionality constant,
shifting the equation from a proportional to an equal relationship by
incorporating known system totals. In some sense, \(k\) serves a role
similar to the gravitational constant in Newton's law -- it calibrates
the model so that outputs match real-world quantities. However, these
real-world quantities are not a set empirical constant (like Newton's
\(G\)) but instead, sensitive to the system and known information about
the system.

From the outset, spatial interaction models emphasized interpretability
of results
\citep{kirbyNormalizingFactorsGravity1970, wilsonSTATISTICALTHEORYSPATIAL1967, wilson1971}.
But unlike earlier approaches that borrowed heuristically from Newtonian
gravity (i.e., interaction between masses over distance), Wilson's
innovation was to generalise the model using \emph{entropy
maximisation}. By maximizing the number of ways individual trip
probabilities could be arranged under known constraints, Wilson derived
models that estimate \emph{statistical averages} of flows between zones
\citep{wilson1971, seniorGravityModellingEntropy1979}.

@@

Additional information can be introduced. For example, when information
is available about the total number of flows produced by each origin,
\(W_i^{(1)}\) in Eq~\ref{eq-phys-gravity-model}, represented as \(O_i\),
then following constraint can be used:

@@

In the framework introduced and inferred from Wilson \citep{wilson1971},
three types of balancing factors are specified: (1) an unconstrained
model that only matches the total volume of interaction \(K\), (2) a
singly constrained model (either by origins \(A_i\) or destinations
\(B_j\) -- explained below), and (3) a doubly constrained model that
satisfies both.

In the unconstrained model, constraints in
Eq~\ref{eq-constraint1-gravitymodel} and
Eq~\ref{eq-constraint2-gravitymodel} do not hold. In practical terms,
this means that the total number of flows predicted by the model must be
equal to sum of all flows from origins \(i\) to destinations \(j\). The
balancing factor \(K\) takes the place of \(k\) and is equal to the
following (as specified in \citep{cliff_evaluating_1974} and
\citep{fotheringham_spatial_1984}):

@@

In the singly constrained model, only constraint
Eq~\ref{eq-constraint1-gravitymodel} or constraint
Eq~\ref{eq-constraint2-gravitymodel} hold. When only
Eq~\ref{eq-constraint1-gravitymodel} holds, entropy maximisation leads
to the production-constrained singly constrained version of
Eq~\ref{eq-phys-gravity-model}, where the proxy for the mass at the
origin \(W_i^{(1)}\) is replaced with \(O_i\). Also, \(k\) is replaced
with a set of balancing factors specific to origins \(A_i\) which
ensures that constraint Eq~\ref{eq-constraint1-gravitymodel} is
satisfied (i.e., the sum of predicted flows from one origin going to all
destinations must equal the known mass at that origin \(O_i\)).
Satisfying this constraint also implicitly fulfills the total constraint
(Eq~\ref{eq-constraint0-gravitymodel}), since the sum of \(O_i\) values
across all origins equals the total number of flows. \(A_i\) takes the
following form:

@@

The singly constrained attraction-constrained model is similar to the
production-constrained version but from the perspective of the mass at
the destination. For the attraction-constrained model, the proxy for the
mass at the destination \(W_j^{(2)}\) is replaced with \(D_j\) in
Eq~\ref{eq-phys-gravity-model}, representing the spatial interaction
inbound flow. Also, \(k\) is replaced with a set of destination-specific
balancing factors \(B_j\) that ensure that
Eq~\ref{eq-constraint2-gravitymodel} is satisfied (hence the total
constraint Eq~\ref{eq-constraint0-gravitymodel} is as well), meaning
that the sum of predicted flows going to one destination from all
origins must equal the known mass of that destination \(D_j\). As
before, destination-specific balancing factors \(B_j\) were derived by
Wilson as:

@@

Lastly, the doubly constrained model is the production-attraction
constrained model in Wilson \citep{wilson1971}. In this case, both
constraints Eq~\ref{eq-constraint1-gravitymodel} and
Eq~\ref{eq-constraint2-gravitymodel} hold simultaneously. These
constraints ensure that the sum of predicted flows from one origin to
all destination, and the predicted flows going to one destination from
all origins must equal the known mass of the origin \(O_i\) and of the
destination \(D_j\). This should hold for all origins and destinations.
The resulting model is, in Wilson's terms, doubly constrained, and from
Eq~\ref{eq-phys-gravity-model}, \(k\) becomes both \(A_i\) and \(B_j\)
shown in Eq~\ref{eq-doubly-constrained-balancing-factors}, and
\(W_i^{(1)}\) and \(W_j^{(2)}\) is replaced with \(O_i\) and \(D_j\).
Derivation of these models is demonstrated in detail elsewhere
\citep[e.g.,][]{ortuzar_2011_modelling, wilsonSTATISTICALTHEORYSPATIAL1967}.

@@

Wilson's work is notable for many reasons. Rather than relying on a
universal constant like \(G\) or a scaling factor to balance units,
Wilson's models calibrate interaction flows through known empirical
constraints using principles of entropy maximisation. This results in
interpretable, balanced (given the system knowns are expressed in the
constraints), and unit-consistent \(ij\) flows. The balancing factors
themselves have been subject to various interpretations -- as terminal
costs \citep{dieter1962distribution}, weighted mean values
\citep{kirbyNormalizingFactorsGravity1970}, or as accessibility measures
themselves \citep{cesario1977new} (as suggested in Wilson
\citep{wilson1971}) and as rents
\citep{morphetGravityModelCalibration2023} -- reflecting ongoing efforts
to understand what these mathematical constructs represent
behaviourally. In this way, the spatial interaction modeling tradition
can be seen to have succeeded where accessibility modeling stalled.
While Wilson's model produces results that are in units of flow tethered
to the system of analysis, which facilitates such interpretations,
Hansen-type measures (still widely used today in accessibility work)
yield outputs that reflect proportional -- not equal -- relationships
and typically lack interpretable units.

Before demonstrating the derivation of family of accessibility measures
that are sensitive to constraints (Section 7), in the next section we
review how conceptually intertwined the spatial interaction modelling
and accessibility literatures are, and where they began to diverge. This
investigation sheds light on why accessibility research may have failed
to adopt a comparable approach, until this paper.

@@

\begin{figure}[H]

\includegraphics[width=0.7\textwidth,height=\textheight]{paper_plos_V3_files/figure-pdf/fig-docs-per-year-1.pdf}

\caption{\label{fig-docs-per-year}Historical pattern of publication:
documents per year.}

\end{figure}%

Despite their close conceptual ties, the accessibility and spatial
interaction modeling literatures have developed along largely separate
paths since the 1970s. Hansen's \citep{hansen1959} formulation of
accessibility became the method used for decades of work on transport
equity, land use analysis, and urban accessibility planning. Meanwhile,
Wilson's \citep{wilson1971} entropy-maximising framework reshaped how
spatial interaction models were constructed, particularly in
transportation demand forecasting. We argue the framework's quiet
innovation -- introducing empirically grounded constraints to shift from
proportionality to calibrated equality -- made the framework immediately
relevant for policy applications as outputs were in tangible units. Yet,
this mechanism was never widely adopted in accessibility analysis.

This divergence is especially striking given the context in which both
frameworks emerged. As noted in
\citep{battyChronicleScientificPlanning1994}, large scale spatial
interaction models (like Wilson's) responded to important developments
at the time, a need ``to meet the dictates and needs of public policy
for strategic land use and transportation planning''. And these needs
were far from trivial: for instance in the U.S., the Federal-Aid Highway
Act of 1956 set in motion the construction of the Interstate Highway
System with an eventual budget exceeding \$600 billion in today's
dollars
\citep{weinerUrbanTransportationPlanning2016, mdotMnDOTJoins2007}. In
this context, spatial interaction models were incorporated into
institutional practices focused on ``predict and provide'' travel demand
forecasting
\citep{kovatch1971modeling, weinerUrbanTransportationPlanning2016}.
Accessibility analysis, by contrast, remained more conceptually diffuse,
focused on indicators of ``potential'' spatial interaction with
opportunities rather than flows that could tangibly guide infrastructure
decisions (e.g., roadway capacity expansion, new construction). Whereas
spatial interaction modelling became a key element of transportation
planning practice, accessibility remained a somewhat more academic
pursuit, and the two streams of literature only rarely connected.

To explore why Wilson's approach may not have crossed over to
accessibility modeling sooner, we conducted a review of the literature
citing Hansen \citep{hansen1959}, Wilson \citep{wilson1971}, or both.
This was done using Web of Science's ``Cited References'' functionality,
and the digital object identifiers of Hansen \citep{hansen1959} and
Wilson \citep{wilson1971}). Only 76 out of the 2,122 documents that
emerged from our search of the General Database cite both. The number of
articles, by year and if they cite Hasen, Wilson, or both are shown in
Fig~\ref{fig-docs-per-year}.

Through the close analysis of why articles that cite both works cite
each work, we identify two distinct patterns: one group of articles
focused on accessibility, the other on spatial interaction. In examining
these groups of articles, we uncover how the relationship between Wilson
and Hansen has often been misunderstood, underexplored, or entirely
overlooked.

In the first stream of literature -- which cite both but are focused
more so on spatial interaction models -- they treat spatial interaction
and accessibility as separate but related phenomenons. Four subsets of
this stream emerge.

@@

Despite their close conceptual ties, accessibility has not meaningfully
absorbed the constraint-based logic of spatial interaction modeling.
This section introduces this conceptual connection by defining a family
of accessibility measures using Wilson's approach grounded in
statistical mechanics, shifting place-based accessibility (\(\grave{a}\)
la Hansen \citep{hansen1959}) from a relationship describing
proportionality to one of equality. This shift addresses the issue of
unit interpretability (associated with the proportional nature of
Hansen-type accessibility indicators previous outlined).

>>>>

\section*{Abstract}
Transportation planning has long prioritized the efficiency of movement.
However, the concept of accessibility represents a more comprehensive
evolution, shifting focus from movement (i.e., trips) to the potential
to spatially interact with desired destinations. Despite growing
recognition of accessibility-based planning approaches, the concept
remains fragmented, with inconsistent definitions and unclear
interpretations. To this end, this paper makes a methodological
contribution by specifying a family of accessibility measures that are
grounded in the shared `gravity-based' theoretical roots of spatial
interaction models, particularly their balancing factors. From this
foundation, we outline four members of the family: the `unconstrained'
measure (i.e., Hansen-type accessibility), the `total-constrained'
measure (i.e., a constrained version of the Hansen-type accessibility),
the `singly-constrained' measure (i.e., related to the popular two-step
floating catchment approach -- 2SFCA), and the `doubly-constrained'
measure representing realized access (i.e., equal to the
doubly-constrained spatial interaction model). These measures can be
interpreted as either the number of accessible opportunities or
accessible population (i.e., market potential). A toy example
illustrates how they produce interpretable unit-based values, offering a
clearer and more coherent basis for accessibility analysis.

\linenumbers

\section{1. Introduction}\label{introduction}

@@

This paper makes two contributions. First, we review how spatial
interaction modeling and accessibility share similar origins but have
diverged in focus and interpretation. Second, we introduce a family of
accessibility measures grounded in spatial interaction principles,
including total, singly, and doubly-constrained cases and variants of
``accessible opportunities'' and ``accessible population''. These cases
and variants align with common measures such as Hansen-type
accessibility \citep{hansen1959}, competitive accessibility measures
such as the 2SFCA method \citep{shen1998, luo2003}, and market potential
models
\citep{harris_market_1954, vickermanAccessibilityAttractionPotential1974}.

@@

The remainder of this paper proceeds as follows. Sections 2 through 4
trace the historical development of spatial interaction and
accessibility research, beginning with Newtonian gravitational analogies
and Carey (1858) \citep{careyPrinciplesSocialScience1858} (Section 2),
moving through early researchers like Ravenstein (1885)
\citep{ravensteinLawsMigration1885} to Stewart (1948)
\citep{stewartDemographicGravitationEvidence1948} who theorized and
formalized spatial interaction patterns (Section 3), and examining how
the ``gravity-based'' accessibility approach in Hansen (1959)
\citep{hansen1959} became the dominant approach in planning practice
(Section 4). Section 5 presents the entropy-based family of spatial
interaction models in Wilson (1971) \citep{wilson1971}, explaining how
the introduction of ``constraints'' (based on top-down known information
as part of entropy-maximization) produces interpretable, unit-consistent
flow estimates. Section 6 explores why accessibility and spatial
interaction literature diverged into separate branches despite their
shared conceptual foundations. Section 7 constitutes the paper's core
contribution: we derive a family of accessibility measures corresponding
to different constraints applied -- unconstrained, singly-constrained,
and doubly-constrained cases -- demonstrating how each yields zone-level
accessibility values expressed in meaningful units (opportunities or
population) rather than unit-inconsistent indices. We illustrate each
measure with numerical examples that clarify the practical implications
of different constraint assumptions. Section 8 concludes by discussing
how these constrained measures can inform planning decisions and improve
clarity in accessibility analysis.

@@

Where \(G\) is the gravitational constant -- an empirically calibrated
value that ensures the model reflects observed forces. Newton's initial
estimate of \(G\) was based on a speculation but received empirical
support after Hutton's and Cavendish's experiments in the late 1700s
\citep{hutton_xxxiii_1778, cavendish_xxi_1798}, which estimated it to
within 1\% accuracy. That is, it took over a century from the
publication of Newton's \emph{Principia Mathematica} to refine the
estimate of the proportionality constant.

@@

Henry C. Carey's \emph{Principles of Social Science}
\citep{careyPrinciplesSocialScience1858} inspired empirical spatial
interaction research in different contexts. Namely, a number of
researchers theoretically and empirically attempted to characterize
human spatial interaction as a force \(F\) directly proportional to the
``masses'' \(M_i\) and \(M_j\) of two locations, and inversely
proportional to their separation distance -- conceptually parallel to
Newtonian gravity, but often omitting a proportionality constant.

@@

What is notable about Eq~\ref{eq-stewart-force}, however, is that the
proportionality constant \(G\) was specified but ``left for future
determination''
\citep[p.~34]{stewartDemographicGravitationEvidence1948}. We can infer
that it is crucial for ensuring \(F\) is maintained in some units of
demographic force.

@@

From Stewart \citep{stewartDemographicGravitationEvidence1948}, we
arrive at 1959 and Walter G. Hansen, whose work proved to be
exceptionally influential in the accessibility literature
\citep{hansen1959}. In this seminal paper, Hansen defined accessibility
as the potential of opportunities for interaction\ldots{} a
generalization of the population-over-distance relationship or
\emph{population potential} concept developed by Stewart
\citep{stewartDemographicGravitationEvidence1948}'' (p.~73). As well as
being a student of city and regional planning at the Massachusetts
Institute of Technology, Hansen was also an engineer with the Bureau of
Roads, and preoccupied with the power of transportation to shape land
uses in a very practical sense. Hansen \citep{hansen1959} drew directly
from Stewart's population potential formula (see
Eq~\ref{eq-stewart-population-potential-sum}), but left aside the
broader (and often problematic) framework of ``social physics''.

@@

\(S_{i}\) in Eq~\ref{eq-accessibility-general} is a measure of the
accessibility from zone \(i\). This is a function of \(O_j\) (the mass
of opportunities at \(j\)), \(d_{ij}\) (the cost, e.g., distance or
travel time, incurred to reach \(j\) from \(i\)), and \(f(d_{ij})\) (a
function that modulates the friction of cost). Today, Hansen is
frequently cited as the father of modern accessibility analysis
\citep[e.g.,][]{reggianiGuestEditorialNew2011}, and Hansen-type
accessibility is commonly referred to as the gravity-based accessibility
measure.

However, Hansen's formulation carried forward a crucial omission that
continues to affect the literature: the proportionality constant \(G\)
included in Stewart's original formulation
(Eq~\ref{eq-stewart-population-potential-sum}) has vanished entirely.
Although Stewart included \(G\) explicitly (with a note that ``\(G\)
{[}was{]} left for future determination: a suitable choice of other
units can reduce it to unity'' {[}p.~34{]}). Hansen made no mention of
it. As a result, modern accessibility analysis has largely evolved
without addressing the constant's role, leaving \(G\) effectively and
implicitly fixed at 1. This omission has significant implications.
Without a proportionality constant, the accessibility formula expresses
only a proportional relationship:
\(S_i \propto \sum_j g(O_j)f(d_{ij})\), not one of calibrated equality.
Recognition of the nature of this relationship is not common in the
literature, but is known, i.e., this proportional equation is shown in
Figure 1 in Wu and Levinson \citep{wuUnifyingAccess2020}.

Furthermore, working with a proportional relationship generates
fundamental issues in comparability between and, arguably, within
studies. Namely, accessibility estimates have no fixed unit, rendering
them sensitive to the choice of impedance functions. For instance, if
travel cost \(d_{ij}\) is measured in meters, then when the travel
impedance function \(f(d_{ij})\) equals \(d_{ij}^{-\beta}\), the
resulting \(S_i\) has units of opportunities per
\(\text{metres}^{\beta}\). However, when \(f(d_{ij})\) is set to equal
\(e^{-\beta d_{ij}}\), the units become opportunities per
\(e^{\beta \text{metres}}\). Such variation impairs comparability across
analyses and obscures the meaning of accessibility scores, making them
difficult to understand and communicate without post-hoc treatment.

Therefore, in practice, Hansen-type measures are ones of proportionality
and are better understood as \emph{ordinal indicators}; they rank
accessibility but lack cardinal meaning or consistent units
\citep{millerAccessibilityMeasurementApplication2018}. The continued
absence of a sort of proportionality constant \(G\) leaves a conceptual
and practical gap in accessibility analysis: a missing link between
theoretical form and empirical measurement that hinders the
interpretability of accessibility measures.

@@

While accessibility research evolved in North America with Hansen
\citep{hansen1959}, a parallel development was taking place across the
Atlantic with Alan G. Wilson. Wilson's groundbreaking paper
\citep{wilson1971} advanced a general framework for spatial interaction
modeling focused on flows of interaction between places and derived from
a non-Newtonian analogy. Additionally, this work was not focused on the
`potential' concept as accessibility had been. Wilson \citep{wilson1971}
formalized the general spatial interaction model through the following
equation:

@@

Of important note for this paper, \(k\) in
Eq~\ref{eq-phys-gravity-model} acts as a proportionality constant,
shifting the equation from a proportional to an equal relationship by
incorporating known system totals. In some sense, \(k\) serves a role
similar to the gravitational constant in Newton's law -- it calibrates
the model so that outputs match real-world quantities. However, these
real-world quantities are not a set empirical constant (like Newton's
\(G\)) but are instead sensitive to the system and known information
about the system.

From the outset, spatial interaction models emphasized interpretability
of results
\citep{kirbyNormalizingFactorsGravity1970, wilsonSTATISTICALTHEORYSPATIAL1967, wilson1971}.
But unlike earlier approaches that borrowed heuristically from Newtonian
gravity (i.e., interaction between masses over distance), Wilson's
innovation was to generalize the model using \emph{entropy
maximization}. By maximizing the number of ways individual trip
probabilities could be arranged under known constraints, Wilson derived
models that estimate \emph{statistical averages} of flows between zones
\citep{wilson1971, seniorGravityModellingEntropy1979}.

@@

Additional information can be introduced. For example, when information
is available about the total number of flows produced by each origin,
\(W_i^{(1)}\) in Eq~\ref{eq-phys-gravity-model}, represented as \(O_i\),
then the following constraint can be used:

@@

In the framework introduced and inferred from Wilson \citep{wilson1971},
three types of balancing factors are specified: (1) an unconstrained
model that only matches the total volume of interaction \(K\), (2) a
singly-constrained model (either by origins \(A_i\) or destinations
\(B_j\) -- explained below), and (3) a doubly-constrained model that
satisfies both.

In the unconstrained model, constraints in
Eq~\ref{eq-constraint1-gravitymodel} and
Eq~\ref{eq-constraint2-gravitymodel} do not hold. In practical terms,
this means that the total number of flows predicted by the model must be
equal to the sum of all flows from origins \(i\) to destinations \(j\).
The balancing factor \(K\) takes the place of \(k\) and is equal to the
following (as specified in \citep{cliff_evaluating_1974} and
\citep{fotheringham_spatial_1984}):

@@

In the singly-constrained model, only constraint
Eq~\ref{eq-constraint1-gravitymodel} or constraint
Eq~\ref{eq-constraint2-gravitymodel} hold. When only
Eq~\ref{eq-constraint1-gravitymodel} holds, entropy maximization leads
to the production-constrained singly-constrained version of
Eq~\ref{eq-phys-gravity-model}, where the proxy for the mass at the
origin \(W_i^{(1)}\) is replaced with \(O_i\). Also, \(k\) is replaced
with a set of balancing factors specific to origins \(A_i\), which
ensures that constraint Eq~\ref{eq-constraint1-gravitymodel} is
satisfied (i.e., the sum of predicted flows from one origin going to all
destinations must equal the known mass at that origin \(O_i\)).
Satisfying this constraint also implicitly fulfills the total constraint
(Eq~\ref{eq-constraint0-gravitymodel}), since the sum of \(O_i\) values
across all origins equals the total number of flows. \(A_i\) takes the
following form:

@@

The singly-constrained attraction-constrained model is similar to the
production-constrained version but from the perspective of the mass at
the destination. For the attraction-constrained model, the proxy for the
mass at the destination \(W_j^{(2)}\) is replaced with \(D_j\) in
Eq~\ref{eq-phys-gravity-model}, representing the spatial interaction
inbound flow. Also, \(k\) is replaced with a set of destination-specific
balancing factors \(B_j\) that ensure that
Eq~\ref{eq-constraint2-gravitymodel} is satisfied (hence the total
constraint Eq~\ref{eq-constraint0-gravitymodel} is as well), meaning
that the sum of predicted flows going to one destination from all
origins must equal the known mass of that destination \(D_j\). As
before, destination-specific balancing factors \(B_j\) were derived by
Wilson as:

@@

Lastly, the doubly-constrained model is the production-attraction
constrained model in Wilson \citep{wilson1971}. In this case, both
constraints Eq~\ref{eq-constraint1-gravitymodel} and
Eq~\ref{eq-constraint2-gravitymodel} hold simultaneously. These
constraints ensure that the sum of predicted flows from one origin to
all destinations, and the predicted flows going to one destination from
all origins, must equal the known mass of the origin \(O_i\) and of the
destination \(D_j\). This should hold for all origins and destinations.
The resulting model is, in Wilson's terms, doubly-constrained, and from
Eq~\ref{eq-phys-gravity-model}, \(k\) becomes both \(A_i\) and \(B_j\)
shown in Eq~\ref{eq-doubly-constrained-balancing-factors}, and
\(W_i^{(1)}\) and \(W_j^{(2)}\) is replaced with \(O_i\) and \(D_j\).
Derivation of these models is demonstrated in detail elsewhere
\citep[e.g.,][]{ortuzar_2011_modelling, wilsonSTATISTICALTHEORYSPATIAL1967}.

@@

Wilson's work is notable for many reasons. Rather than relying on a
universal constant like \(G\) or a scaling factor to balance units,
Wilson's models calibrate interaction flows through known empirical
constraints using principles of entropy maximization. This results in
interpretable, balanced (given the system knowns are expressed in the
constraints), and unit-consistent \(ij\) flows. The balancing factors
themselves have been subject to various interpretations -- as terminal
costs \citep{dieter1962distribution}, weighted mean values
\citep{kirbyNormalizingFactorsGravity1970}, or as accessibility measures
themselves \citep{cesario1977new} (as suggested in Wilson
\citep{wilson1971}) and as rents
\citep{morphetGravityModelCalibration2023} -- reflecting ongoing efforts
to understand what these mathematical constructs represent
behaviourally. In this way, the spatial interaction modeling tradition
can be seen to have succeeded where accessibility modeling stalled.
While Wilson's model produces results that are in units of flow tethered
to the system of analysis, which facilitates such interpretations,
Hansen-type measures (still widely used today in accessibility work)
yield outputs that reflect proportional -- not equal -- relationships
and typically lack interpretable units.

Before demonstrating the derivation of the family of accessibility
measures that are sensitive to constraints (Section 7), in the next
section, we review how conceptually intertwined the spatial interaction
modelling and accessibility literatures are, and where they began to
diverge. This investigation sheds light on why accessibility research
may have failed to adopt a comparable approach until this paper.

@@

Despite their close conceptual ties, the accessibility and spatial
interaction modeling literatures have developed along largely separate
paths since the 1970s. Hansen's \citep{hansen1959} formulation of
accessibility became the method used for decades of work on transport
equity, land use analysis, and urban accessibility planning. Meanwhile,
Wilson's \citep{wilson1971} entropy-maximizing framework reshaped how
spatial interaction models were constructed, particularly in
transportation demand forecasting. We argue the framework's quiet
innovation -- introducing empirically grounded constraints to shift from
proportionality to calibrated equality -- made the framework immediately
relevant for policy applications as outputs were in tangible units. Yet,
this mechanism was never widely adopted in accessibility analysis.

This divergence is especially striking given the context in which both
frameworks emerged. As noted in
\citep{battyChronicleScientificPlanning1994}, large-scale spatial
interaction models (like Wilson's) responded to important developments
at the time, a need ``to meet the dictates and needs of public policy
for strategic land use and transportation planning''. And these needs
were far from trivial: for instance, in the U.S., the Federal-Aid
Highway Act of 1956 set in motion the construction of the Interstate
Highway System with an eventual budget exceeding \$600 billion in
today's dollars
\citep{weinerUrbanTransportationPlanning2016, mdotMnDOTJoins2007}. In
this context, spatial interaction models were incorporated into
institutional practices focused on ``predict and provide'' travel demand
forecasting
\citep{kovatch1971modeling, weinerUrbanTransportationPlanning2016}.
Accessibility analysis, by contrast, remained more conceptually diffuse,
focused on indicators of ``potential'' spatial interaction with
opportunities rather than flows that could tangibly guide infrastructure
decisions (e.g., roadway capacity expansion, new construction). Whereas
spatial interaction modelling became a key element of transportation
planning practice, accessibility remained a somewhat more academic
pursuit, and the two streams of literature only rarely connected.

To explore why Wilson's approach may not have crossed over to
accessibility modeling sooner, we conducted a review of the literature
citing Hansen \citep{hansen1959}, Wilson \citep{wilson1971}, or both.
This was done using Web of Science's ``Cited References'' functionality,
and the digital object identifiers of Hansen \citep{hansen1959} and
Wilson \citep{wilson1971}. Only 76 out of the 2,122 documents that
emerged from our search of the General Database cite both. The number of
articles, by year and if they cite Hasen, Wilson, or both, are shown in
Figure 1.

Through the close analysis of why articles that cite both works cite
each work, we identify two distinct patterns: one group of articles
focused on accessibility, the other on spatial interaction. In examining
these groups of articles, we uncover how the relationship between Wilson
and Hansen has often been misunderstood, under-explored, or entirely
overlooked.

In the first stream of literature -- which cite both but are focused
more on spatial interaction models -- they treat spatial interaction and
accessibility as separate but related phenomena. Four subsets of this
stream emerge.

@@

Despite their close conceptual ties, accessibility has not meaningfully
absorbed the constraint-based logic of spatial interaction modeling.
This section introduces this conceptual connection by defining a family
of accessibility measures using Wilson's approach grounded in
statistical mechanics, shifting place-based accessibility (\(\grave{a}\)
la Hansen \citep{hansen1959}) from a relationship describing
proportionality to one of equality. This shift addresses the issue of
unit interpretability (associated with the proportional nature of
Hansen-type accessibility indicators previously outlined).
